---
title: "Classification of exercise acivity using the weight lifting dataset from"
author: "zchen"
date: "February 27, 2016"
output: html_document
---

```{r, echo=F}
library(caret);
library(ggplot2); library(grid); library(gridExtra)
options(digits=5)
```

## Summary

The goal of this analysis is to model and predict the quality of execution of weight lifting exercises
using data from accelerometers attached to the human subject’s body and dumbbell. First the dataset was
filtered by by removing the variables consisting mostly of missing values and/or with zero or near zero
variance. Then the training set was splited into internal training and testing sets (70:30 ratio). The
internal training set was used to train and built a list of models, including linear discriminant
analysis (LDA), quadratic discriminant analysis (QDA), tree based model (RPART), random forest (RF),
tree based boosting (GBM), model based boosting (MBOOST) using the Caret package with 5 fold cross
validation method (no repeats). The performance of these models was evaluated using the internal testing
dataset. It shows that the random forest model is the best performing model, giving a prediction
accuracy of over 99%, thus this modle was used to predict on the external testing data with 20
observations.


## Background and Dataset

Human Activity Recognition (HAR) has gained increasing attention by the computing research community
in recent years, due to the development of context-aware systems. There are many potential applications
for HAR, e.g., elderly monitoring, life log systems for monitoring energy expenditure and for
supporting weight-loss programs, and digital assistants for weight lifting exercises, etc.

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har.
Six young health participants were asked to perform one set of 10 repetitions of the Unilateral
Dumbbell Biceps Curl in five different fashions:

Class A: exactly according to the specification  
Class B: throwing the elbows to the front  
Class C: lifting the dumbbell only halfway  
Class D: lowering the dumbbell only halfway  
Class E: throwing the hips to the front.  

In this project, the goal is to use data from accelerometers on the belt, forearm, arm, and dumbell
of 6 participants, to predict the five classes of activities.

The training data and test data for this project are downloaded from:  
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv  
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

More information about the data can be found at: http://groupware.les.inf.puc-rio.br/har.


##Load data and exploratory analysis

The training and testing datasets were downloaded from above links. The training data set has 19622
observations and 160 columns, and the testing data set has 20 obersvations and same number of columns.

```{r, echo=F}
#load training and testing data
tr <- read.csv("pml-training.csv")
te <- read.csv("pml-testing.csv")
dim(tr); dim(te)
```

First, the training set was filtered by dropping zero or near zero variance variables using nearZeroVar
function from the Caret package, and columns with largely missing values (NA) are also removed. After
filtering, the remaining training set has 19622 observations and 53 variables.

```{r, echo=F}
# filtering data by dropping zero or near zero variance variables
# columns with mostly missing values (NA) are also removed
tr <- tr[,-c(1:7)]
nz <- nearZeroVar(tr)
tr <- tr[, -nz]
nac <-sapply(tr, function(y) sum(length(which(is.na(y)))))
trf <- tr[,!names(tr) %in% names(nac[nac==19216])]
dim(trf)
```

*Scatter plots between several measurements, colored based on different classes*

```{r, echo=F}
q1 <- qplot(x=trf$accel_forearm_x, y=trf$accel_forearm_z, color=trf$classe)
q2 <- qplot(x=trf$accel_forearm_y, y=trf$accel_forearm_z, color=trf$classe)
q3 <- qplot(x=trf$accel_dumbbell_x, y=trf$accel_dumbbell_z, color=trf$classe)
q4 <- qplot(x=trf$accel_dumbbell_y, y=trf$accel_dumbbell_z, color=trf$classe)
grid.arrange(q1, q2, q3, q4, ncol=4)
```


## Data preprocessing and modelling

The training dataset was splitted into an interanl training set and internal testing set (ratio 70:30),
so we can build the models using the same internal training set and measure the performance of all the
models using same internal testing set, thus we can have an unbiased way to evaluate the models.

We tested a variety of models, including linear discriminant analysis (LDA), quadratic discriminant
analysis (QDA), tree based model (RPART), random forest (RF), tree based boosting (GBM), model based
boosting (MBOOST) using the Caret package with 5 fold cross validation method (no repeats). The
performance of these models was evaluated using the same internal testing set.


```{r, echo=F}
set.seed(333)
intr <- createDataPartition(y=trf$classe, p=0.7, list=FALSE)
cvTr <- trf[intr, ]
cvTe <- trf[-intr, ]
trCtr <- trainControl(method="cv", number=5)

m_lda <- train(classe ~ ., data=cvTr, method="lda", trControl=trCtr)
m_qda <- train(classe ~ ., data=cvTr, method="qda", trControl=trCtr)

m_tree <- train(classe ~ ., data=cvTr, method="rpart", trControl=trCtr)
#mf <- train(classe~.,data=cvTr, method="rf", trControl= trainControl(method="cv", number=5, allowParallel=TRUE))
m_rf <- train(classe ~ ., data=cvTr, method="rf", trControl=trCtr, allowParallel=TRUE, prox=TRUE)
m_gbm <- train(classe ~ ., data=cvTr, method="gbm", trControl=trCtr, verbose=F, allowParallel=TRUE)
m_mboost <- train(classe ~ ., data=cvTr, method="mboost", trControl=trCtr, verbose=F, allowParallel=TRUE)
```


## Model performance evaluation

To estimate the accuracy and the out of sample error of the models, the internal testing dataset with
was used. The observations in this set were not used in model generation and all have a known classe assignment, which was used for comparison with the predicted one.

```{r, echo=F}
pform_lda <- confusionMatrix(predict(m_lda, cvTe), cvTe$classe)
pform_qda <- confusionMatrix(predict(m_qda, cvTe), cvTe$classe)
pform_tree <- confusionMatrix(predict(m_tree, cvTe), cvTe$classe)
pform_rf <- confusionMatrix(predict(m_rf, cvTe), cvTe$classe)
pform_gbm <- confusionMatrix(predict(m_gbm, cvTe), cvTe$classe)
pform_mboost <- confusionMatrix(predict(m_mboost, cvTe), cvTe$classe)

res <- cbind(pform_lda$overall, pform_qda$overall, pform_tree$overall, pform_gbm$overall, pform_mboost$overall, pform_rf$overall)
names(res) <- c("LDA", "QDA", "RPART", "GBM", "MBOOST", "RF")
res
```


*The result show that RF model is the best performing model.*


*predict the external testing set with 20 observations using RF model*

```{r, echo=F}
#predict the external testing set with 20 observations using RF model
pred2Submit <- predict(m_rf, te[names(te) %in% names(cvTr)])
pred2Submit
```


## Conclusions

A list of models were built using the Caret package to predict different classies of activities, the results show that the best performing model is random forest model, which achieves over 99% of accuracy with internal testing set and  100% accuracy on the external testing data set.



